---
title: "Analysis: Dog Poop Complaints in Manhattan"
format: 
  html:
    code-fold: true
---

```{r}
#| label: setup
#| include: true
#| warning: false
#| message: false

# Use quarto render to generate a static site in /docs
# Set github pages to deploy from main branch, /docs

library(flextable)
library(sf)
library(tidyverse)
library(socratadata)
library(lubridate)
library(mapgl)
library(glue)

# If needed, ensure that the Quarto render and RStudio read the same environment vars
readRenviron(".Renviron")
```

New York City's 311 program has six service request categories related to dog poop. They are:

```{r, message=FALSE}
suppressMessages(dog_poop_sr <- read_csv("dog_poop_SRs.csv"))

dog_poop_sr |> 
  flextable() |> 
  set_table_properties(width = 1, layout = "autofit") |> 
  align(j = 1, align = "left", part = "all")
```

```{r}
dog_poop_complaints <- soc_read(
  "https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2020-to-Present/erm2-nwe9/about_data",
  query = soc_query(
    where = "borough = 'MANHATTAN' AND ((descriptor = 'Dog Waste') OR (descriptor = 'Animal Waste' AND descriptor_2 = 'Dog'))"
  )
)

min_date <- min(as_date(dog_poop_complaints$created_date))
max_date <- max(as_date(dog_poop_complaints$created_date))
n_complaints <- nrow(dog_poop_complaints)

```

Based on these six service request types, there were **`r scales::comma(n_complaints)`** dog poop-related service requests filed in Manhattan between **`r min_date`** and **`r max_date`**.

A clear annual trend is visible when aggregating these service requests to year and month:
```{r}
#| out-width: 100%


monthly_dog_poop_complaints <-  dog_poop_complaints |> 
  st_drop_geometry() |> 
  group_by(yearMonth = paste0(
    year(created_date),
                  "-",
                  str_split_i(date(created_date),
                              "-",
                              2))) |> 
  summarize(monthly_complaints = n()) |> 
  ungroup() |> 
  # Pin to first of month for plotting
  mutate(date = as.Date(paste0(yearMonth, "-01")))

ggplot(monthly_dog_poop_complaints,
       aes(x = date,
           y = monthly_complaints)) +
  geom_point()+
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "1 month",
    date_labels = "%Y"
  ) +
  labs(x = "Date (Year, Month)",
       y = "Dog Poop Service Requests")
```

Since 2022, dog poop-related 311 service requests have tended to be highest in January, February, and March, with a decreasing trend throughout the calendar year. 2024 is an exception to this trend; the count of service requests increased throughout the calendar year. 

Based on the peak winter months in 2022, 2023, and 2025, the overall volume of dog-poop service requests in Manhattan appears to be increasing.

The back-to-back snowstorms in February of 2026 appear to have led to approximately three times the number of dog poop related service requests than in February of 2025.

```{r}
dog_poop_complaints_2026 <- dog_poop_complaints |> 
  filter(year(created_date)>2025)

dog_poop_complaints_2026_by_cb <- dog_poop_complaints |> 
  st_drop_geometry() |> 
  filter(year(created_date)>2025) |> 
  group_by(community_board) |> 
  summarize(complaints_2026 = n())
```

In 2026 alone, there have been **`r nrow(dog_poop_complaints_2026)`** dog poop related service requests. The spatial distribution of these complaints is not homogenous:

```{r}
dog_poop_complaints_2026_by_cb |> 
  flextable() |> 
  set_table_properties(width = 1, layout = "autofit") |> 
  align(j = 1, align = "left", part = "all") |> 
  set_header_labels(
    community_board = "Community Board",
    complaints_2026 = "Dog Poop Service Requests"
  )
```

In 2026, the greatest number of dog poop related service requests have come from Manhattan Community Districts 10 and 12. Interestingly, a few Community Districts (1, 2, 5) have had very few in comparison.

Visually:
```{r}
suppressMessages(
comDist <- read_sf("https://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/NYC_Community_Districts/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=pgeojson") |> 
  st_make_valid() |> 
  filter(BoroCD < 200) |> 
  mutate(community_board = paste0(
    str_sub(as.character(BoroCD), -2),
    " MANHATTAN"
  )) |> 
  left_join(dog_poop_complaints_2026_by_cb) |> 
  mutate(
    tooltip_html = glue(
      "<b>Community Board:</b> {community_board}<br>",
      "<b>Service requests:</b> {complaints_2026}"
    )
  )
)

maplibre(style = carto_style("positron")) |> 
  fit_bounds(comDist) |> 
  add_fill_layer(
    id = "Community Districts",
    source = comDist,
    fill_color = interpolate(
      column = "complaints_2026",
      values = c(
        min(comDist$complaints_2026),
        max(comDist$complaints_2026)
      ),
      stops = c("#fff5f0", "#ef3b2c")
    ),
    fill_opacity = 0.7,
    tooltip = "tooltip_html"
  ) |> 
  add_circle_layer(
    id = "Service Requests",
    source = dog_poop_complaints_2026,
    circle_color = "#7f7f7f",
    circle_stroke_color = "#cccccc",
    circle_radius = 5,
    circle_stroke_width = 1
  )
```

