---
title: "Analysis: Dog Poop Complaints in Manhattan"
format: 
  html:
    code-fold: true
---

```{r}
#| label: setup
#| include: true

# Use quarto render to generate a static site in /docs
# Set github pages to deploy from main branch, /docs

library(flextable)
library(sf)
library(tidyverse)
library(socratadata)
library(lubridate)
library(mapgl)
library(glue)
library(scales)
library(ggrepel)

# If needed, ensure that the Quarto render and RStudio read the same environment vars
readRenviron(".Renviron")
```

Per its data dictionary on [NYC Open Data](https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2020-to-Present/erm2-nwe9/about_data), New York City's 311 program includes **5,963** specific Service Request (SR) types that can be directed to specific agencies. Of these nearly 6,000 ways to formally complain to the city, exactly six relate to dog poop. They are:

```{r}
#| cache: true


dog_poop_sr <- read_csv("dog_poop_SRs.csv")

dog_poop_complaints <- soc_read(
  "https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2020-to-Present/erm2-nwe9/about_data",
  query = soc_query(
    where = "borough = 'MANHATTAN' AND ((descriptor = 'Dog Waste') OR (descriptor = 'Animal Waste' AND descriptor_2 = 'Dog'))"
  )
)

mn_SRs_20_26_countOnly <- soc_read(
  "https://data.cityofnewyork.us/resource/erm2-nwe9.json",
  query = soc_query(
    select = "count(*) as count",
    where = "borough = 'MANHATTAN'"
  )
)

min_date <- min(as_date(dog_poop_complaints$created_date))
max_date <- max(as_date(dog_poop_complaints$created_date))
n_complaints <- nrow(dog_poop_complaints)

dog_poop_sr |> 
  flextable() |> 
  set_table_properties(width = 1, layout = "autofit") |> 
  align(j = 1, align = "left", part = "all")
```

Between **`r min_date`** and **`r max_date`**, there were **`r scales::comma(n_complaints)`** dog poop-related service requests filed in Manhattan. Dog poop-related service requests accounted for **`r round(n_complaints/mn_SRs_20_26_countOnly[[1,1]][1] *100,2)`%** of all service requests filed in Manhattan for the same time period. 

A clear annual trend is visible when aggregating these service requests to year and month:
```{r}
#| out-width: 100%


monthly_dog_poop_complaints <-  dog_poop_complaints |> 
  st_drop_geometry() |> 
  group_by(yearMonth = paste0(
    year(created_date),
                  "-",
                  str_split_i(date(created_date),
                              "-",
                              2))) |> 
  summarize(monthly_complaints = n()) |> 
  ungroup() |> 
  # Pin to first of month for plotting
  mutate(date = as.Date(paste0(yearMonth, "-01")))

ggplot() +
  geom_point(data = monthly_dog_poop_complaints,
       aes(x = date,
           y = monthly_complaints))+
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "1 month",
    date_labels = "%Y"
  ) +
  labs(x = "Date (Year, Month)",
       y = "Dog Poop Service Requests")
```

Since 2022, dog poop-related 311 service requests have tended to be highest in January, February, and March, with a decreasing trend throughout the calendar year. 2024 is an exception to this trend; the count of service requests increased throughout the calendar year. 

Based on the peak winter months in 2022, 2023, and 2025, the overall volume of dog-poop service requests in Manhattan appears to be increasing.

The back-to-back snowstorms in February of 2026 appear to have led to approximately three times the number of dog poop related service requests than in February of 2025.

An annual trend is less apparent when considering Manhattan Service Requests of all types during the same time period:

```{r}
#| cache: true
#| out-width: 100%


mn_complaints_byMonthYear <- soc_read(
  "https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2020-to-Present/erm2-nwe9/about_data",
  query = soc_query(
    select = "date_trunc_ym(created_date) AS year_month, count(*) AS monthly_complaints",
    where  = "borough = 'MANHATTAN'",
    group  = "date_trunc_ym(created_date)",
    order  = "year_month"
  )
)

ggplot() +
  geom_point(data = mn_complaints_byMonthYear,
             aes(x = as.Date(year_month),
                 y = monthly_complaints)) +
  scale_x_date(
    date_breaks = "1 year",
    date_minor_breaks = "1 month",
    date_labels = "%Y"
  ) +
  scale_y_continuous(
    label = comma
  ) +
  labs(x = "Date (Year, Month)",
       y = "All Service Requests")
```
<br />
When considering the spatial variation in a particular Service Request type as it relates to quality of life concerns at a given geographic unit (here, Community Districts are used), two factors must be considered: 

- 311 Service Request data are records of citizen interaction with a government system, not direct observation of the associated phenomenon, and
- There is spatial variation in 311 contacting volume, which is a function of contacting propensity and condition. See Minkoff (2016)[^1] for further discussion.

[^1]: [Minkoff, Scott L. "NYC 311: A tract-level analysis of citizen–government contacting in New York City." Urban Affairs Review 52.2 (2016): 211-246.](https://doi.org/10.1177/1078087415577796)

Therefore, when considering the variation in dog poop-related Service Requests across Manhattan Community Districts, it must be reiterated that this is not an actual measure of dog poop on Manhattan streets. We also expect that Service Request volume will largely account for the variation in dog poop Service Requests. Areas where this trend doesn't hold warrant further investigation. 

A linear model of dog poop service requests as a function of service request volume at the Community District level allows for an evaluation of the above relationship. Model residuals (i.e. the error, or how different the observed relationship between dog poop service requests and request volume is from the expected relationship) are a useful measure of which Community Districts have greater or fewer dog poop service requests than expected.

```{r}
#| cache: true

mn_SRs_20_26 <- soc_read(
  "https://data.cityofnewyork.us/resource/erm2-nwe9.json",
  query = soc_query(
    select = "community_board, count(*) as count",
    where = "borough = 'MANHATTAN'",
    group_by = "community_board"
  )
)

mn_dp_SRs_20_26 <- soc_read(
  "https://data.cityofnewyork.us/resource/erm2-nwe9.json",
  query = soc_query(
    select = "community_board, count(*) as dp_count",
    where = "borough = 'MANHATTAN' AND ((descriptor = 'Dog Waste') OR (descriptor = 'Animal Waste' AND descriptor_2 = 'Dog'))",
    group_by = "community_board"
  )
)
```

```{r}
#| out-width: 100%


mn_comparison_20_26 <- mn_dp_SRs_20_26 |> 
  left_join(mn_SRs_20_26) |> 
  filter(community_board != "Unspecified MANHATTAN")

fit <- lm(dp_count ~ count, data = mn_comparison_20_26)

ci <- predict(fit, interval = "confidence", level = 0.95)
mn_comparison_20_26$lwr <- ci[, "lwr"]
mn_comparison_20_26$upr <- ci[, "upr"]
mn_comparison_20_26$outlier <- with(mn_comparison_20_26,
                                     dp_count < lwr | dp_count > upr)

ggplot(mn_comparison_20_26,
       aes(x = count, y = dp_count)) +
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  geom_text_repel(
    data = subset(mn_comparison_20_26, outlier),
    aes(label = community_board),   # swap in whatever your CB label column is
    size = 3,
    min.segment.length = 0,
    box.padding = 0.4
  ) +
  scale_x_continuous(labels = label_comma()) +
  labs(x = "Service Requests, All Categories",
       y = "Dog Poop Service Requests",
       title = "Manhattan 311 Dog Poop Service Requests as a Function of Request Volume",
       subtitle = "2020 - 2026 (YTD), by Community District")+
  scale_y_continuous(minor_breaks = seq(-250,700,50))
```

The linear regression of dog poop-related service requests as a function of overall request volume in Manhattan Community Boards (2020 - 2026 YTD) is statistically significant with a p-value of less than **0.01.** The adjusted *R^2^* of the regression is **0.6957**, meaning that **69.57%** of the variance in dog poop service requests in Manhattan is accounted for by request volume. Those Community Districts whose observed service request volumes are of interest. Manhattan Community Districts 2, 3, and 5 have lower-than-expected counts of dog poop service requests since January 1, 2020, while only Manhattan Community District 12 has a higher than expected count of dog poop service requests. 

Given the basis for identifying Community District 12 as an area of concern, we can look for those blocks which have the highest number of requests as potential sites for further investigation:

```{r}
#| cache: true


comDist12 <- read_sf("https://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/NYC_Community_Districts/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=pgeojson") |> 
  st_make_valid() |> 
  filter(BoroCD == 112) |> 
  mutate(community_board = paste0(
    str_sub(as.character(BoroCD), -2),
    " MANHATTAN"
  ))

comDist12_wkt <- st_as_text(st_geometry(comDist12)[[1]])

comDist12_dog_poop_SRs <- soc_read(
  "https://data.cityofnewyork.us/resource/erm2-nwe9.json",
  query = soc_query(
    where = paste0(
      "borough = 'MANHATTAN' AND community_board = '12 MANHATTAN' AND ((descriptor = 'Dog Waste') OR (descriptor = 'Animal Waste' AND descriptor_2 = 'Dog'))"
      )
    )
  )

comDist12_centerlines <- soc_read(
  "https://data.cityofnewyork.us/City-Government/Centerline/inkn-q76z.json",
  query = soc_query(
    where = paste0(
      "within_polygon(the_geom, '",
      comDist12_wkt,
      "')"
    )
  )
)
```

Each of the **`r nrow(comDist12_dog_poop_SRs)`** dog poop-related Service Requests filed in Manhattan Community District 12 between 2020 and 2026 (YTD) can be associated with the nearest block per the Street Centerline dataset. The blocks with the **10** highest counts of nearest Service Requests are shown below. 

```{r}
#| out-width: 100%
#| out-height: 1000px


comDist12_dog_poop_SRs <- comDist12_dog_poop_SRs |> 
  st_join(
    comDist12_centerlines |> select(objectid),
    join = st_nearest_feature
    )

top_blocks <- comDist12_dog_poop_SRs |> 
  group_by(objectid) |> 
  summarize(n_dog_poop_SRs = n()) |> 
  arrange(desc(n_dog_poop_SRs)) |> 
  head(10)

maplibre(style = carto_style("positron")) |> 
  fit_bounds(comDist12) |> 
  add_line_layer(
    id = "Community District 12",
    source = comDist12,
    line_color = "#888888",
    line_width = 2
  ) |>
   add_line_layer(
  id = "Top Blocks Hitbox",
  source = top_blocks |>
    mutate(
      tooltip_html = glue(
        "<b>Block ObjectID:</b> {objectid}<br>",
        "<b>Dog Poop Service Requests</b> {n_dog_poop_SRs}"
      )
    ),
  tooltip = "tooltip_html",
  line_color = "#000000",
  line_width = 24,
  line_opacity = 0
  ) |>
  add_line_layer(
    id = "Top Blocks",
    source = comDist12_centerlines |> 
      filter(objectid %in% top_blocks$objectid),
    line_color = "#b30000",
    line_width = 8
  ) |> 
  add_circle_layer(
    id = "Service Requests",
    source = comDist12_dog_poop_SRs |> 
      filter(objectid %in% top_blocks$objectid),
    circle_color = "#444444",
    circle_stroke_color = "#666666",
    circle_radius = 10,
    circle_stroke_width = 2,
    cluster_options = cluster_options(
      max_zoom = 20
    )
  )
```
<br />
The top 10 blocks (street segments, to be specific) represent **`r round(sum(top_blocks$n_dog_poop_SRs)/nrow(comDist12_dog_poop_SRs)*100,2)`%** of the dog poop-related Service Requests. Notably, there is a cluster of **360** Service Requests located on and around Riverside Drive between West 158th and West 160th streets.

Based on the above, accurate, appropriately contextualized, and defensible messaging may be something like:

> We crunched the numbers -- Since 2020, New Yorkers filed 80 to 360 more dog poop-related Service Requests in Manhattan Community District 12 than expected based on request volume. We decided to take a closer look. Over 60% of the service requests were filed on and around Riverside Drive between West 158th and West 160th streets. Other clusters of requests came from the West side of Hood Wright Park, Fort Washington Ave between West 187th and West 190th streets, near Hillside Ave and Borgardus Place, and on Seaman Ave between Dyckman and and Cumming Streets.

Having established Manhattan Community District 12 as a potential area of concern, let's turn our attention to hot spots of dog poop service requests in all of Manhattan's 12 community districts. 

In doing so, we'll limit the time frame to **January 1, 2024 to 2026 year to date**

```{r}
#| cache: true

manhattan_comDist <- read_sf("https://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/NYC_Community_Districts/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=pgeojson") |> 
  st_make_valid() |> 
  filter(BoroCD < 200) |> 
  mutate(community_board = paste0(
    str_sub(as.character(BoroCD), -2),
    " MANHATTAN"
  ))

mn_wkt <- st_as_text(st_geometry(st_union(manhattan_comDist))[[1]])

mn_dp_SRs_24_26 <- soc_read(
  "https://data.cityofnewyork.us/resource/erm2-nwe9.json",
  query = soc_query(
    where = "borough = 'MANHATTAN'
             AND created_date >= '2024-01-01T00:00:00'
             AND ((descriptor = 'Dog Waste') OR (descriptor = 'Animal Waste' AND descriptor_2 = 'Dog'))
    "
  )
)

mn_centerlines <- soc_read(
  "https://data.cityofnewyork.us/City-Government/Centerline/inkn-q76z.json",
  query = soc_query(
    where = "boroughcode = '1'"
  )
) 

mn_dp_SRs_24_26 <- mn_dp_SRs_24_26 |> 
  st_join(
    mn_centerlines |> select(objectid),
    join = st_nearest_feature
  )

top_blocks_manhattan <- mn_dp_SRs_24_26 |> 
  group_by(community_board, objectid) |> 
  summarise(n_dog_poop_SRs = n()) |> 
  arrange(desc(n_dog_poop_SRs)) |> 
  slice(1:3) |> 
  filter(!(community_board %in% c("64 MANHATTAN", "Unspecified MANHATTAN", "08 BRONX")))

top_blocks_manhattan_counts <- top_blocks_manhattan |> 
  st_drop_geometry() |> 
  filter(community_board != "12 MANHATTAN") |> 
  pull(n_dog_poop_SRs)

waste_bag_dispensers <- soc_read(
  "https://data.cityofnewyork.us/Recreation/Canine-Waste-Dispensers/5npv-j6gn.json",
  query = soc_query(
    where = "borough = 'M'"
  )
)

mn_litter_basket <- soc_read(
  "https://data.cityofnewyork.us/Environment/DSNY-Litter-Basket-Inventory/8znf-7b2c.json",
  query = soc_query(
    where = paste0(
      "within_polygon(point, '",
      mn_wkt,
      "')"
    )
  )
)

mn_pops <- soc_read(
  "https://data.cityofnewyork.us/City-Government/Privately-Owned-Public-Spaces-POPS-/rvih-nhyn.json",
  query = soc_query(
    where = "borough_name = 'Manhattan'"
  )
)

mn_dog_runs <- soc_read(
  "https://data.cityofnewyork.us/Recreation/Dog-Runs/hxx3-bwgv.json",
  query = soc_query(
    where = paste0(
      "within_polygon(the_geom, '",
      mn_wkt,
      "')"
    )
  )
)
```
There were **`r scales::comma(nrow(mn_dp_SRs_24_26))`** dog poop-related Service Requests filed in Manhattan From January 1, 2024 to April 19, 2026. Similar to the map above, each service request was associated with the nearest city block. The blocks with the **three** highest counts of service requests are shown below. Aside from Community District 12, the range of highest service request counts per block is low, ranging from **`r max(top_blocks_manhattan_counts[seq(1,31,3)])`** to **`r min(top_blocks_manhattan_counts[seq(1,31,3)])`**. These blocks are shown highlighted in dark red and represent hot spots of dog-poop service requests per Community District.

The locations of the **`r nrow(waste_bag_dispensers)`** dog waste bag dispensers in Manhattan are shown below in dark green. The majority of the dispensers are located in Community District 3. The small number of these dispensers makes it difficult to discern any trends with respect to Service Requests. With **`r scales::comma(nrow(mn_litter_basket))`** locations in Manhattan, DSNY-served litter baskets are much more prevalent than dog waste bag dispensers. The locations are shown in light green below, and are hidden by default. Use the layer control in the top right corner to turn the layer on. Litter baskets in NYC Parks or Privately-Owned Public spaces (POPs) are not shown due to a lack of data availability. The location of POPs may also be turned on using the map's layer control. 


```{r}
#| out-width: 100%
#| out-height: 1000px

maplibre(style = carto_style("positron")) |> 
  fit_bounds(manhattan_comDist) |> 
  add_line_layer(
    id = "Manhattan Community Districts",
    source = manhattan_comDist,
    line_color = "#888888",
    line_width = 1
  ) |> 
  add_fill_layer(
    id = "Dog Runs",
    source = mn_dog_runs,
    fill_color = "#ff83fb"
  ) |> 
  add_line_layer(
    id = "Top Blocks",
    source = mn_centerlines |> 
      filter(objectid %in% top_blocks_manhattan$objectid),
    line_color = "#b30000",
    line_width = 8
  ) |> 
  add_circle_layer(
    id = "pops",
    source = mn_pops,
    circle_color = "orange",
    circle_radius = 5,
    visibility = "none"
  ) |> 
  add_circle_layer(
    id = "Litter Baskets",
    source = mn_litter_basket,
    circle_color = "lightgreen",
    circle_radius = 5,
    visibility = "none"
    #min_zoom = 16
  )|> 
  add_circle_layer(
    id = "Service Requests",
    source = mn_dp_SRs_24_26 |> 
      filter(objectid %in% top_blocks_manhattan$objectid),
    circle_color = "#444444",
    circle_stroke_color = "#666666",
    circle_radius = 10,
    circle_stroke_width = 2,
    cluster_options = cluster_options(
      max_zoom = 14.5
    )
  ) |> 
  add_circle_layer(
    id = "Canine Waste Bag Dispensers",
    source = waste_bag_dispensers,
    circle_color = "green",
    circle_radius = 5
  ) |> 
  add_layers_control(
    layers = list(
      "Litter Baskets" = "Litter Baskets",
      "POPs" = "pops"
    ),
    position = "top-right"
  ) |> 
  add_categorical_legend(
    legend_title = "Layers",
    values = c("Canine Waste Bag Dispensers",
               "Litter Baskets",
               "POPs", "Dog Runs"),
    colors = c("green", "lightgreen", "orange", "#ff83fb"),
    patch_shape = "circle",
    style = list(
      background_opacity = 0.8
    )
  )
```
<br />
In general, litter baskets can be found at the intersections of most city streets. However, there are a few notable exceptions. Rather than just displaying clusters of dog poop Service Requests, the below maps display all dog poop Service requests filed between January 1, 2024 and April 19, 2026.

In Community District 12, there are relatively few litter baskets in the area immediately surrounding the cluster of dog poop service requests:
```{r}
#| out-width: 100%
#| out-height: 700px

maplibre(style = carto_style("positron"),
         center = c(-73.94676835,40.83643540),
         zoom = 16) |> 
  add_line_layer(
    id = "Manhattan Community Districts",
    source = manhattan_comDist,
    line_color = "#888888",
    line_width = 1
  ) |> 
  add_fill_layer(
    id = "Dog Runs",
    source = mn_dog_runs,
    fill_color = "#ff83fb"
  ) |> 
  add_circle_layer(
    id = "pops",
    source = mn_pops,
    circle_color = "orange",
    circle_radius = 5,
    visibility = "visible"
  ) |> 
  add_circle_layer(
    id = "Litter Baskets",
    source = mn_litter_basket,
    circle_color = "lightgreen",
    circle_radius = 5,
    visibility = "visible"
    #min_zoom = 16
  )|> 
  add_circle_layer(
    id = "Service Requests",
    source = mn_dp_SRs_24_26,
    circle_color = "#b30000",
    circle_radius = 10,
    cluster_options = cluster_options(
      max_zoom = 15.5
    )
  ) |> 
  add_circle_layer(
    id = "Canine Waste Bag Dispensers",
    source = waste_bag_dispensers,
    circle_color = "green",
    circle_radius = 5
  ) |> 
  add_layers_control(
    layers = list(
      "Litter Baskets" = "Litter Baskets",
      "POPs" = "pops"
    ),
    position = "top-right"
  ) |> 
  add_categorical_legend(
    legend_title = "Layers",
    values = c("Dog Poop Service Requests","Canine Waste Bag Dispensers",
               "Litter Baskets",
               "POPs", "Dog Runs"),
    colors = c("#b30000","green", "lightgreen", "orange", "#ff83fb"),
    patch_shape = "circle",
    style = list(
      background_opacity = 0.8
    )
  )
```
<br />
The same issue is seen in Community District 10 on West 142nd and 5th Ave:
```{r}
#| out-width: 100%
#| out-height: 700px

maplibre(style = carto_style("positron"),
         center = c(-73.936006, 40.817830),
         zoom = 15.5) |> 
  add_line_layer(
    id = "Manhattan Community Districts",
    source = manhattan_comDist,
    line_color = "#888888",
    line_width = 1
  ) |> 
  add_fill_layer(
    id = "Dog Runs",
    source = mn_dog_runs,
    fill_color = "#ff83fb"
  ) |> 
  add_circle_layer(
    id = "pops",
    source = mn_pops,
    circle_color = "orange",
    circle_radius = 5,
    visibility = "visible"
  ) |> 
  add_circle_layer(
    id = "Litter Baskets",
    source = mn_litter_basket,
    circle_color = "lightgreen",
    circle_radius = 5,
    visibility = "visible"
    #min_zoom = 16
  )|> 
  add_circle_layer(
    id = "Service Requests",
    source = mn_dp_SRs_24_26,
    circle_color = "#b30000",
    circle_radius = 10,
    cluster_options = cluster_options(
      max_zoom = 15.5
    )
  ) |> 
  add_circle_layer(
    id = "Canine Waste Bag Dispensers",
    source = waste_bag_dispensers,
    circle_color = "green",
    circle_radius = 5
  ) |> 
  add_layers_control(
    layers = list(
      "Litter Baskets" = "Litter Baskets",
      "POPs" = "pops"
    ),
    position = "top-right"
  ) |> 
  add_categorical_legend(
    legend_title = "Layers",
    values = c("Dog Poop Service Requests","Canine Waste Bag Dispensers",
               "Litter Baskets",
               "POPs", "Dog Runs"),
    colors = c("#b30000","green", "lightgreen", "orange", "#ff83fb"),
    patch_shape = "circle",
    style = list(
      background_opacity = 0.8
    )
  )
```
<br />
There's only one litter basket north of Central Park on Adam Clayton Powell Jr. Boulevard south of West 115th Street:

```{r}
#| out-width: 100%
#| out-height: 700px

maplibre(style = carto_style("positron"),
         center = c(-73.95443343, 40.80068412),
         zoom = 15.5) |> 
  add_line_layer(
    id = "Manhattan Community Districts",
    source = manhattan_comDist,
    line_color = "#888888",
    line_width = 1
  ) |> 
  add_fill_layer(
    id = "Dog Runs",
    source = mn_dog_runs,
    fill_color = "#ff83fb"
  ) |> 
  add_circle_layer(
    id = "pops",
    source = mn_pops,
    circle_color = "orange",
    circle_radius = 5,
    visibility = "visible"
  ) |> 
  add_circle_layer(
    id = "Litter Baskets",
    source = mn_litter_basket,
    circle_color = "lightgreen",
    circle_radius = 5,
    visibility = "visible"
    #min_zoom = 16
  )|> 
  add_circle_layer(
    id = "Service Requests",
    source = mn_dp_SRs_24_26,
    circle_color = "#b30000",
    circle_radius = 10,
    cluster_options = cluster_options(
      max_zoom = 15.5
    )
  ) |> 
  add_circle_layer(
    id = "Canine Waste Bag Dispensers",
    source = waste_bag_dispensers,
    circle_color = "green",
    circle_radius = 5
  ) |> 
  add_layers_control(
    layers = list(
      "Litter Baskets" = "Litter Baskets",
      "POPs" = "pops"
    ),
    position = "top-right"
  ) |> 
  add_categorical_legend(
    legend_title = "Layers",
    values = c("Dog Poop Service Requests","Canine Waste Bag Dispensers",
               "Litter Baskets",
               "POPs", "Dog Runs"),
    colors = c("#b30000","green", "lightgreen", "orange", "#ff83fb"),
    patch_shape = "circle",
    style = list(
      background_opacity = 0.8
    )
  )
```
<br />
There's less litter basket density around the Tudor City hotspot than in surrounding areas:

```{r}
#| out-width: 100%
#| out-height: 700px

maplibre(style = carto_style("positron"),
         center = c(-73.97300990, 40.74680816),
         zoom = 15.5) |> 
  add_line_layer(
    id = "Manhattan Community Districts",
    source = manhattan_comDist,
    line_color = "#888888",
    line_width = 1
  ) |> 
  add_fill_layer(
    id = "Dog Runs",
    source = mn_dog_runs,
    fill_color = "#ff83fb"
  ) |> 
  add_circle_layer(
    id = "pops",
    source = mn_pops,
    circle_color = "orange",
    circle_radius = 5,
    visibility = "visible"
  ) |> 
  add_circle_layer(
    id = "Litter Baskets",
    source = mn_litter_basket,
    circle_color = "lightgreen",
    circle_radius = 5,
    visibility = "visible"
    #min_zoom = 16
  )|> 
  add_circle_layer(
    id = "Service Requests",
    source = mn_dp_SRs_24_26,
    circle_color = "#b30000",
    circle_radius = 10,
    cluster_options = cluster_options(
      max_zoom = 15.5
    )
  ) |> 
  add_circle_layer(
    id = "Canine Waste Bag Dispensers",
    source = waste_bag_dispensers,
    circle_color = "green",
    circle_radius = 5
  ) |> 
  add_layers_control(
    layers = list(
      "Litter Baskets" = "Litter Baskets",
      "POPs" = "pops"
    ),
    position = "top-right"
  ) |> 
  add_categorical_legend(
    legend_title = "Layers",
    values = c("Dog Poop Service Requests","Canine Waste Bag Dispensers",
               "Litter Baskets",
               "POPs", "Dog Runs"),
    colors = c("#b30000","green", "lightgreen", "orange", "#ff83fb"),
    patch_shape = "circle",
    style = list(
      background_opacity = 0.8
    )
  )
```
<br />
The same is true for the cluster south of 1 Centre St: 

```{r}
#| out-width: 100%
#| out-height: 700px

maplibre(style = carto_style("positron"),
         center = c(-74.005017, 40.710385),
         zoom = 16) |> 
  add_line_layer(
    id = "Manhattan Community Districts",
    source = manhattan_comDist,
    line_color = "#888888",
    line_width = 1
  ) |> 
  add_fill_layer(
    id = "Dog Runs",
    source = mn_dog_runs,
    fill_color = "#ff83fb"
  ) |> 
  add_circle_layer(
    id = "pops",
    source = mn_pops,
    circle_color = "orange",
    circle_radius = 5,
    visibility = "visible"
  ) |> 
  add_circle_layer(
    id = "Litter Baskets",
    source = mn_litter_basket,
    circle_color = "lightgreen",
    circle_radius = 5,
    visibility = "visible"
    #min_zoom = 16
  )|> 
  add_circle_layer(
    id = "Service Requests",
    source = mn_dp_SRs_24_26,
    circle_color = "#b30000",
    circle_radius = 10,
    cluster_options = cluster_options(
      max_zoom = 15.5
    )
  ) |> 
  add_circle_layer(
    id = "Canine Waste Bag Dispensers",
    source = waste_bag_dispensers,
    circle_color = "green",
    circle_radius = 5
  ) |> 
  add_layers_control(
    layers = list(
      "Litter Baskets" = "Litter Baskets",
      "POPs" = "pops"
    ),
    position = "top-right"
  ) |> 
  add_categorical_legend(
    legend_title = "Layers",
    values = c("Dog Poop Service Requests","Canine Waste Bag Dispensers",
               "Litter Baskets",
               "POPs", "Dog Runs"),
    colors = c("#b30000","green", "lightgreen", "orange", "#ff83fb"),
    patch_shape = "circle",
    style = list(
      background_opacity = 0.8
    )
  )
```
<br />
In the quest to address stubborn quality of life issues in Manhattan, a further exploration of litter basket deserts may be of interest as the call to action (i.e. the installation of more baskets) is straightforward.

